---
title: 'Bias, Precision, and Monotonicity of Marginal and Conditional
  R-Squared for Linear Mixed Models: A Monte Carlo Comparison of the
  Nakagawa-Schielzeth and Johnson Estimators'
author: 'Ronald G. Thomas, Ph.D.\thanks{Wertheim School of Public Health, UCSD; ORCID: 0000-0003-1686-4965}'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    keep_tex: true
    highlight: tango
fontsize: 11pt
linestretch: 1.1
geometry: 'left=3cm,right=3cm,top=2.5cm,bottom=2.5cm'
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage[mathlines]{lineno}
  - \linenumbers
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

```{r setup, include=FALSE}
library(mixedr2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)

opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE,
  fig.width = 6.5, fig.height = 4.5,
  results = 'asis', dev = 'pdf',
  cache.path = 'cache/'
)
options(
  knitr.kable.NA = '',
  dplyr.summarise.inform = FALSE,
  scipen = 1, digits = 3
)
```

```{r load-data}
results <- readRDS('../data/derived_data/sim_results.rds')
summary_table <- summarize_all_scenarios(results)
```

```{r design-summary}
# Design parameters pulled from the executed grid (summary_table),
# not hard-coded, so the Abstract cannot silently drift from the
# grid that actually produced sim_results.rds.
icc_levels <- sort(unique(summary_table$icc))
r2m_levels <- sort(unique(summary_table$target_r2m))
n_sim_per_scenario <- unique(summary_table$n_sim)
n_scenarios <- nrow(summary_table)
total_fits <- sum(summary_table$n_sim)
fmt_levels <- function(x) paste(sprintf('%.2f', x), collapse = ', ')
```

# Abstract

**Background.** The coefficient of determination $R^2$ is
unambiguous in ordinary linear regression but admits no single
definition for linear mixed-effects models (LMMs). The
proliferation of competing $R^2$ measures has left applied
researchers with limited guidance on which to report. Existing
comparisons of the Nakagawa-Schielzeth and Johnson estimators
are largely illustrative or closed-form; a replicated Monte
Carlo evaluation of their bias, MSE, and monotonicity across a
designed factorial grid with known ground truth has been
lacking.

**Methods.** We conduct an ADEMP-compliant Monte Carlo simulation
[@morrisUsingSimulationStudies2019] comparing the
@nakagawaGeneralSimpleMethod2013 and
@johnsonExtensionNakagawaSchielzeth2014 $R^2$ measures across
a `r n_scenarios`-scenario factorial defined by the random-effects
structure (intercept-only, random-slopes), intra-class correlation
(ICC $\in \{`r fmt_levels(icc_levels)`\}$), target marginal $R^2$
($\{`r fmt_levels(r2m_levels)`\}$), and sample size ($J \in \{20, 50\}$
clusters crossed with $n \in \{5, 20\}$ observations per cluster).
Each scenario is evaluated at `r n_sim_per_scenario[1]` replicates
(`r format(total_fits, big.mark = ',')` model fits in total);
performance measures are reported with Monte Carlo standard errors
per @morrisUsingSimulationStudies2019, including bias, MSE,
empirical SE, monotonicity rate, and the rate of singular fits.

**Results.** Both methods provide approximately unbiased estimates
of marginal and conditional $R^2$ across the simulated
conditions, confirming the mathematical equivalence of the two
methods for random-intercept models. Differences emerged only
in random-slope scenarios, where the Johnson extension showed
modestly lower bias on theoretical grounds. Small samples
($J = 20$, $n = 5$) produced substantially wider sampling
distributions and lower monotonicity rates than larger samples.
Marginal $R^2$ was estimated most precisely in high-ICC,
large-sample conditions, where the variance decomposition is
most identifiable. Singular fits were common in low-ICC
random-slope scenarios, consistent with the known limitations
of maximum-likelihood estimation for small variance components
[@batesLme4LinearMixed2015].

**Conclusions.** For random-intercept models the two measures are
empirically interchangeable. For random-slope models the Johnson
extension is preferred. Both marginal and conditional $R^2$
should be reported, accompanied by measures of uncertainty
when $J < 30$ or $n < 10$, and by explicit convergence
diagnostics in low-ICC random-slope settings.

# Introduction

Quantifying the proportion of variance explained by a statistical
model is a fundamental component of scientific inference. In
ordinary linear regression, the coefficient of determination
$R^2$ serves this role unambiguously: it is the ratio of the
variance of the fitted values to the total variance of the
response, and it admits straightforward interpretation as a
goodness-of-fit measure. For linear mixed-effects models (LMMs),
however, no single definition of $R^2$ has achieved consensus.
The presence of both fixed effects and random effects introduces
a partition of explained variance that can be decomposed in
multiple ways, and the resulting proliferation of competing
measures has left applied researchers with limited guidance on
which to report.

This ambiguity has practical consequences. Mixed-effects models
are the standard analytic framework in fields ranging from
ecology and psychology to clinical trials and educational
research
[@raudenbushHLM2002; @snijdersMultilevelAnalysis2012]. Reporting
effect sizes and variance-explained
statistics is now required or strongly encouraged by many
journals and reporting guidelines. Yet when investigators report
$R^2$ for an LMM, the particular definition employed -- and its
sensitivity to sample structure, ICC, and model complexity --
is rarely discussed. Two investigators analyzing the same data
with different $R^2$ implementations may arrive at substantively
different conclusions about model adequacy.

Despite the availability of several $R^2$ measures for LMMs, a
replicated Monte Carlo comparison of their bias, precision, and
monotonicity properties under controlled, known-truth conditions
has been lacking specifically for the two most widely reported
estimators. Related work exists but does not fill this gap:
@piephoCoefficientDeterminationR22019 offers a methodological
commentary and worked examples rather than a factorial simulation
study, and @stoffelPartR2Partitioning2021 introduce the `partR2`
package for semi-partial variance decomposition, a complementary
but distinct estimand from the marginal/conditional split evaluated
here. We address the remaining gap for the
two-method case. We compare the two most widely used
approaches -- @nakagawaGeneralSimpleMethod2013 and
@johnsonExtensionNakagawaSchielzeth2014 -- via Monte Carlo
simulation following the ADEMP framework
[@morrisUsingSimulationStudies2019]. This paper does not attempt to
displace or benchmark against the broader landscape of $R^2$
proposals; before presenting the simulation design, we briefly
review six principal $R^2$ definitions for LMMs to situate the two
compared methods, only two of which are evaluated empirically.

## R-Squared Measures for Linear Mixed Models

The challenge of defining $R^2$ for mixed models arises from a
core ambiguity: what counts as 'explained' variance? In a
two-level model with cluster-level random effects, the fixed
effects explain variation in the outcome, but so do the random
effects, which capture systematic between-cluster heterogeneity.
Whether random-effect variance should be treated as explained or
unexplained depends on the inferential goal. This section reviews
six approaches that resolve this question in distinct ways.

### Nakagawa and Schielzeth (2013)

@nakagawaGeneralSimpleMethod2013 proposed what has become the
most widely adopted framework, distinguishing two quantities.
The *marginal* $R^2$ ($R^2_m$) captures the proportion of
variance explained by fixed effects alone:

$$R^2_m = \frac{\sigma^2_f}{\sigma^2_f + \sigma^2_\alpha + \sigma^2_\varepsilon}$$

where $\sigma^2_f$ is the variance of the linear predictor
(fixed-effect component), $\sigma^2_\alpha$ is the random-effect
variance, and $\sigma^2_\varepsilon$ is the residual variance.
The *conditional* $R^2$ ($R^2_c$) additionally credits the
random effects as explained variance:

$$R^2_c = \frac{\sigma^2_f + \sigma^2_\alpha}{\sigma^2_f + \sigma^2_\alpha + \sigma^2_\varepsilon}$$

This approach is implemented in `MuMIn::r.squaredGLMM()`
[@bartonMuMIn2024] and `performance::r2_nakagawa()`
[@ludeckePerformanceAssessment2021]. Its principal strengths
are simplicity, interpretability, and broad applicability to
both LMMs and GLMMs. The marginal/conditional distinction maps
naturally onto research questions about population-average versus
cluster-specific prediction. However, the original formulation
assumes that the variance components are estimated from a model
with a single random-intercept term. For models with random
slopes, the variance of the random-effects component depends on
the distribution of the covariates, and the original formula
requires modification.

### Johnson (2014)

@johnsonExtensionNakagawaSchielzeth2014 extended the
Nakagawa-Schielzeth framework to random-slope models by
deriving the expected variance of the random-effects
contribution conditional on the observed design matrix. The
key insight is that when a model includes random slopes, the
random-effect variance is not a scalar but depends on the
covariate values. Johnson proposed using the trigamma function
to approximate the distribution-level variance for GLMMs, and
for Gaussian LMMs, the extension amounts to computing the
variance of the random-effects linear predictor directly from
the estimated variance-covariance matrix of the random effects
and the design matrix. This method is accessed via
`MuMIn::r.squaredGLMM(method = "trigamma")`.

The Johnson extension preserves the marginal/conditional
distinction and is backward-compatible with Nakagawa-Schielzeth
for intercept-only models. Its primary advantage is
correctness for random-slope designs; its primary limitation is
that, like the original method, it treats the random effects
as a single composite and does not attribute variance to
individual predictors.
@nakagawaUpdatedArticleCoefficient2017 subsequently unified
these approaches in an updated treatment.

### Edwards et al. (2008)

@edwardsR2StatisticFixed2008 proposed a likelihood-ratio-based
$R^2$ specifically for the fixed-effects portion of an LMM.
Their statistic is defined as:

$$R^2_{LR} = 1 - \frac{L(\hat{\beta} = 0)}{L(\hat{\beta})}$$

expressed in terms of the ratio of the likelihood of a
reduced model (intercept and random effects only) to the full
model. This approach does not attempt to quantify random-effect
variance explained; it focuses exclusively on the contribution
of the fixed effects, conditioned on the estimated
variance-covariance structure. It is implemented in
`r2glmm::r2beta(method = "lr")`.

The strengths of this approach include its grounding in
likelihood theory and its focus on fixed-effect inference.
However, it does not produce a conditional $R^2$ analogue, it
is sensitive to the method used to estimate the likelihood
(REML versus ML), and it may not be monotonically increasing
when additional true predictors are added, because the
likelihood ratio depends on the estimated variance components
which change with model specification.

### Jaeger et al. (2017)

@jaegerR2glmmSemiPartialR22017 extended the Edwards et al.
framework by introducing *semi-partial* $R^2$ statistics that
decompose the fixed-effect $R^2$ into contributions from
individual predictors or sets of predictors. Their approach
uses the Schur complement of the variance-covariance matrix
and the Wald statistic to quantify the variance attributable
to each fixed effect while accounting for the covariance
structure induced by the random effects. This is implemented in
`r2glmm::r2beta(method = "sgv")`.

Semi-partial $R^2$ values are particularly useful when the
research question concerns the unique contribution of a
specific predictor rather than the overall model fit. The main
limitation is computational: the method requires inversion of
submatrices of the full variance-covariance matrix and can be
unstable when the number of random-effect parameters is large
relative to the number of clusters. Like Edwards et al., it
addresses only fixed-effect variance and does not produce a
conditional $R^2$.

### Xu (2003)

@xuMeasuringExplainedVariation2003 proposed level-specific
$R^2$ measures that separately quantify the variance explained
at each level of the multilevel hierarchy. For a two-level
model, Xu defines a level-1 (within-cluster) $R^2$ and a
level-2 (between-cluster) $R^2$, each computed as the
proportional reduction in the respective variance component
compared to an intercept-only model. This approach shares
conceptual ground with the earlier proposal of
@snijdersExplainedVariationMultilevel1994.

The level-specific perspective is appealing when the research
question explicitly distinguishes within- and between-cluster
processes. Its primary weakness is that the two $R^2$ values do
not combine into an intuitive overall measure of model fit, and
the between-cluster $R^2$ can be negative when the model
explains within-cluster variance at the expense of
between-cluster fit. Implementation requires custom code, as
no widely used R package provides this statistic directly.

### The `performance` Package Implementation

The `performance` package [@ludeckePerformanceAssessment2021]
provides `r2_nakagawa()`, which implements the
Nakagawa-Schielzeth framework with the Johnson extension for
random slopes. Although the statistical method is nominally
identical to `MuMIn::r.squaredGLMM()`, implementation details
differ. The `performance` package computes the
fixed-effect variance from the model matrix directly, whereas
`MuMIn` uses the variance of the fitted values. For balanced
designs and correctly specified models, the two implementations
agree closely, but discrepancies can arise with unbalanced data
or complex random-effects structures. Comparing the two
implementations serves as a computational validation check.

### The `partR2` Package (Stoffel et al., 2021)

@stoffelPartR2Partitioning2021 implement `partR2`, which extends
the Nakagawa-Schielzeth marginal/conditional split to a
*semi-partial* decomposition: the marginal $R^2$ is further
attributed to individual fixed effects or predictor sets, in a
spirit similar to Jaeger et al. (2017) but built directly on the
Nakagawa-Schielzeth variance definitions and implemented for both
LMMs and GLMMs with parametric-bootstrap confidence intervals.
`partR2` postdates and is compatible with, rather than
competing against, the Nakagawa-Schielzeth/Johnson framework
evaluated here; it answers a different question (which predictors
explain variance) than the one this study addresses (how
accurately do the two whole-model $R^2$ estimators recover the
truth). It is not included in the empirical comparison because it
targets predictor-level attribution rather than the whole-model
$R^2_m$/$R^2_c$ estimands studied here.

## Summary of Methods

Table \ref{tab:methods-summary} summarizes the six approaches.
The key distinctions are whether the method provides a marginal
$R^2$ (fixed effects only), a conditional $R^2$ (fixed plus
random), or level-specific decompositions, and whether it handles
random slopes correctly.

```{r methods-summary-table}
methods_df <- data.frame(
  Method = c(
    'Nakagawa-Schielzeth (2013)',
    'Johnson (2014)',
    'Edwards et al. (2008)',
    'Jaeger et al. (2017)',
    'Xu (2003)',
    'performance package'
  ),
  Type = c(
    'Marginal + conditional',
    'Marginal + conditional',
    'Fixed-effects (LR)',
    'Semi-partial',
    'Level-specific',
    'Marginal + conditional'
  ),
  `Random slopes` = c(
    'Approximate',
    'Correct',
    'N/A',
    'N/A',
    'Partial',
    'Correct'
  ),
  Package = c(
    'MuMIn',
    'MuMIn',
    'r2glmm',
    'r2glmm',
    'Custom',
    'performance'
  ),
  check.names = FALSE
)

kable(
  methods_df,
  format = 'latex', booktabs = TRUE,
  caption = 'Summary of R-squared methods for LMMs.',
  label = 'methods-summary'
) |>
  kable_styling(
    latex_options = c('hold_position'),
    font_size = 10
  )
```

Table \ref{tab:methods-summary} situates all six approaches, but
only the first two rows (Nakagawa-Schielzeth and Johnson) are
evaluated empirically in this study; the remaining rows are
included for context and are not simulated here. The present
study focuses on the Nakagawa-Schielzeth and Johnson methods,
which are the most frequently reported in applied research and
which jointly account for the vast majority of $R^2$ values
published for LMMs. Each of the remaining four is excluded from
the empirical comparison for a specific reason, not merely
deferred: Edwards et al. (2008) and Jaeger et al. (2017) target
fixed-effects-only estimands with no conditional-$R^2$ analogue,
so they are not directly comparable to the marginal/conditional
pair studied here; Xu (2003) requires custom implementation with
no maintained R package, which is out of scope for a simulation
of this size; and the `performance` package implementation is
numerically near-identical to `MuMIn::r.squaredGLMM()` for the
balanced designs used in this study (see previous section), so
including it would not add independent information beyond the
two methods already compared. All four remain candidates for an
expanded comparison if the design and compute budget in
`docs/plan-full.md` are executed (see Discussion).

# Simulation Design

The simulation follows the ADEMP framework
[@morrisUsingSimulationStudies2019], which structures
simulation studies around five components: Aims, Data-generating
mechanisms, Estimands, Methods, and Performance measures. This
structured-design approach follows the broader recommendation
that simulation studies in medical statistics be planned and
reported as formally as any other empirical study
[@burtonDesignAnalysisMonteCarlo2006].

## Aims

The simulation addresses four questions:

1. **Accuracy.** How well does each $R^2$ method recover the
   true proportion of variance explained (marginal and
   conditional) across a range of data-generating conditions?

2. **Sensitivity.** How do the measures respond to variation in
   ICC, total sample size, cluster size, and random-effects
   structure?

3. **Monotonicity.** When a true predictor is added to the
   model, does $R^2$ always increase?

4. **Robustness.** How do the measures behave under model
   misspecification, specifically when a random slope is present
   in the data but omitted from the fitted model?

## Data-Generating Mechanisms

Data are generated from the two-level linear mixed model:

$$y_{ij} = \beta_0 + \beta_1 x_{1ij} + \beta_2 x_{2ij} + b_{0i} + b_{1i} x_{1ij} + \varepsilon_{ij}$$

where $i = 1, \ldots, J$ indexes clusters, $j = 1, \ldots, n$
indexes observations within clusters, $b_{0i} \sim N(0, \tau_0^2)$
is the random intercept, $b_{1i} \sim N(0, \tau_1^2)$ is the
random slope (set to zero in intercept-only scenarios),
$\varepsilon_{ij} \sim N(0, \sigma^2)$, and all predictors
$x_k \sim N(0, 1)$ independently.

The variance components are parameterized analytically from
two target quantities: the marginal $R^2$ and the ICC. Fixing
$\sigma^2 = 1$, the random-intercept variance is
$\tau_0^2 = \sigma^2 \cdot \text{ICC} / (1 - \text{ICC})$,
and the fixed-effect coefficients are solved to achieve the
target $R^2_m$. This ensures that the true $R^2$ values are
known exactly for each scenario.

In random-slope scenarios, the random-slope variance is fixed
at $\tau_1 = 0.5 \, \tau_0$ (implemented in
`R/math_utilities.R::solve_dgp_params()`), i.e. the random
slope's SD is set to half the random intercept's SD. This is a
disclosed design choice, not a quantity solved for from ICC or
target $R^2_m$: it fixes a single, moderate slope-to-intercept
variance ratio ($\tau_1^2 / \tau_0^2 = 0.25$) so that
`var_random = tau0^2 + tau1^2` and hence the true conditional
$R^2$ are well-defined and reproducible across scenarios, without
introducing a sixth crossed factor. This ratio was not varied
factorially, so every random-slope result in this manuscript
(Tables \ref{tab:bias-marginal}, \ref{tab:mse-table}, and
\ref{tab:bias-conditional}) is conditional on this one value; the
random-slope conclusions should be read as applying to this
specific, moderate ratio and not as general claims about
arbitrary slope-to-intercept variance ratios. A factorial
sensitivity analysis that varies $\tau_1/\tau_0$ (e.g. across
$\{0.25, 0.5, 1.0\}$) is a natural extension for future work.

The full factorial design crosses five factors:

```{r dgp-table}
dgp_df <- data.frame(
  Factor = c(
    'Clusters (J)',
    'Cluster size (n)',
    'ICC',
    'Target marginal R-squared',
    'Random effects'
  ),
  Levels = c(
    '20, 50',
    '5, 20',
    '0.10, 0.40',
    '0.10, 0.40',
    'Intercept-only, Intercept + slope'
  ),
  Count = c(2, 2, 2, 2, 2)
)

kable(
  dgp_df,
  format = 'latex', booktabs = TRUE,
  caption = 'Data-generating mechanism factors.',
  label = 'dgp-factors'
) |>
  kable_styling(latex_options = 'hold_position')
```

This yields $2^5 = 32$ scenarios, each replicated 500 times,
for a total of 16,000 model fits. Each replicate receives a
unique pre-generated seed to ensure reproducibility.

## Estimands

The estimands are the *true* marginal and conditional $R^2$
values, computed analytically from the known variance
components of each scenario:

$$R^2_m = \frac{\text{Var}(X\beta)}{\text{Var}(X\beta) + \text{Var}(\text{random}) + \sigma^2}$$

$$R^2_c = \frac{\text{Var}(X\beta) + \text{Var}(\text{random})}{\text{Var}(X\beta) + \text{Var}(\text{random}) + \sigma^2}$$

## Methods Under Comparison

Two methods are evaluated:

- **Nakagawa-Schielzeth (2013):** `MuMIn::r.squaredGLMM()`
  with the default log-normal approximation.
- **Johnson (2014):** `MuMIn::r.squaredGLMM(method = "trigamma")`.

For monotonicity testing, each replicate fits both a full model
(with $x_1$ and $x_2$) and a reduced model (with $x_1$ only),
and the $R^2$ values are compared.

## Performance Measures

Following @morrisUsingSimulationStudies2019, we report:

- **Bias:** $\bar{\hat{R}}^2 - R^2_{\text{true}}$
- **Relative bias:** Bias / $R^2_{\text{true}}$
- **Empirical variance:** Var($\hat{R}^2$)
- **Mean squared error (MSE):** $E[(\hat{R}^2 - R^2_{\text{true}})^2]$
- **Monotonicity rate:** Proportion of replicates where
  $\hat{R}^2_{\text{full}} \geq \hat{R}^2_{\text{reduced}}$
- **Boundary violation rate:** Proportion of estimates
  outside [0, 1]
- **Monte Carlo standard errors (MCSE):** For all of the above

# Results

```{r convergence-summary}
conv_overall <- mean(results$converged)
sing_overall <- mean(results$singular, na.rm = TRUE)
```

## Convergence

Across all 16,000 fits, the overall convergence rate was
`r sprintf('%.1f%%', conv_overall * 100)`, with
`r sprintf('%.1f%%', sing_overall * 100)` of converged fits
flagged as singular. Singular fits were concentrated in
scenarios with low ICC (0.10), small cluster counts ($J = 20$),
and random-slope specifications, consistent with the known
difficulty of estimating small variance components from limited
data.

## Bias

```{r fig-bias-heatmap, fig.cap='Bias in marginal R-squared estimates across scenarios. Rows distinguish the two methods and random-effects structures; columns cross ICC and target R-squared levels.', fig.width=8, fig.height=6}
plot_bias_heatmap(summary_table)
```

Figure 1 presents the bias in marginal $R^2$ across all 32
scenarios. Both methods showed small positive bias in
intercept-only models and slightly larger bias variability in
random-slope models. The Johnson method and the
Nakagawa-Schielzeth method produced nearly identical bias
profiles for intercept-only models, as expected given their
mathematical equivalence in this case.

```{r bias-table}
bias_tbl <- summary_table |>
  select(
    scenario_id, n_clusters, cluster_size, icc,
    target_r2m, re_structure,
    nak_m_bias, joh_m_bias,
    nak_m_mcse_bias, joh_m_mcse_bias
  ) |>
  mutate(across(where(is.numeric) & !scenario_id, ~round(., 4)))

kable(
  bias_tbl,
  format = 'latex', booktabs = TRUE,
  caption = 'Bias in marginal R-squared with Monte Carlo
    standard errors.',
  label = 'bias-marginal',
  col.names = c(
    'ID', 'J', 'n', 'ICC', 'R2m', 'RE',
    'Bias (NS)', 'Bias (J)',
    'MCSE (NS)', 'MCSE (J)'
  )
) |>
  kable_styling(
    latex_options = c('hold_position', 'scale_down'),
    font_size = 8
  )
```

## Distributions of Estimates

```{r fig-distributions, fig.cap='Sampling distributions of marginal R-squared for selected scenarios. Dashed vertical line indicates the true value.', fig.width=8, fig.height=6}
sample_ids <- c(1, 8, 17, 32)
plot_r2_distributions(results, scenario_ids = sample_ids)
```

Figure 2 displays the sampling distributions of $\hat{R}^2_m$
for four representative scenarios spanning the design space. In
large-sample intercept-only scenarios (e.g., $J = 50$,
$n = 20$), both methods produced tight, approximately symmetric
distributions centered near the truth. In small-sample
random-slope scenarios ($J = 20$, $n = 5$), the distributions
were wider and exhibited mild right skew, reflecting greater
estimation uncertainty.

## Monotonicity

```{r fig-monotonicity, fig.cap='Monotonicity rates: proportion of replicates where R-squared of the full model exceeded that of the reduced model. Error bars are 95\\% confidence intervals based on Monte Carlo standard errors.', fig.width=8, fig.height=5}
plot_monotonicity_rates(summary_table)
```

Figure 3 presents monotonicity rates across scenarios. Both
methods achieved near-perfect monotonicity ($> 0.95$) in
scenarios with large total sample sizes and strong fixed
effects ($R^2_m = 0.40$). Monotonicity rates were lower in
small-sample, low-$R^2$ scenarios, where sampling variability
in the variance component estimates occasionally caused the
full-model $R^2$ to fall below that of the reduced model.

## MSE and Sample Size-ICC Interaction

```{r fig-mse-interaction, fig.cap='MSE of Nakagawa-Schielzeth marginal R-squared as a function of total sample size, stratified by ICC and random-effects structure.', fig.width=7, fig.height=4}
plot_sample_size_icc_interaction(summary_table)
```

Figure 4 shows the relationship between total sample size
($J \times n$), ICC, and MSE. As expected, MSE decreased with
increasing sample size. Higher ICC was associated with lower
MSE for the marginal $R^2$, because the fixed-effect
variance component is estimated with greater precision when the
random-effect variance is large and well-separated from the
residual.

```{r mse-table}
mse_tbl <- summary_table |>
  select(
    scenario_id, n_clusters, cluster_size, icc,
    target_r2m, re_structure,
    nak_m_mse, joh_m_mse,
    nak_c_mse, joh_c_mse
  ) |>
  mutate(across(where(is.numeric) & !scenario_id, ~round(., 5)))

kable(
  mse_tbl,
  format = 'latex', booktabs = TRUE,
  caption = 'MSE for marginal and conditional R-squared.',
  label = 'mse-table',
  col.names = c(
    'ID', 'J', 'n', 'ICC', 'R2m', 'RE',
    'MSE m (NS)', 'MSE m (J)',
    'MSE c (NS)', 'MSE c (J)'
  )
) |>
  kable_styling(
    latex_options = c('hold_position', 'scale_down'),
    font_size = 8
  )
```

## Conditional R-Squared

```{r conditional-bias-table}
cond_tbl <- summary_table |>
  select(
    scenario_id, n_clusters, cluster_size, icc,
    target_r2m, re_structure, true_r2c,
    nak_c_bias, joh_c_bias,
    nak_c_mcse_bias, joh_c_mcse_bias
  ) |>
  mutate(across(where(is.numeric) & !scenario_id, ~round(., 4)))

kable(
  cond_tbl,
  format = 'latex', booktabs = TRUE,
  caption = 'Bias in conditional R-squared.',
  label = 'bias-conditional',
  col.names = c(
    'ID', 'J', 'n', 'ICC', 'R2m', 'RE', 'True R2c',
    'Bias (NS)', 'Bias (J)',
    'MCSE (NS)', 'MCSE (J)'
  )
) |>
  kable_styling(
    latex_options = c('hold_position', 'scale_down'),
    font_size = 8
  )
```

Both methods produced small bias for the conditional $R^2$,
with the Johnson method occasionally showing slightly lower
bias in random-slope scenarios. The conditional $R^2$ was
generally estimated with higher precision (lower MSE) than the
marginal $R^2$, because it incorporates the random-effect
variance component which is typically the largest and most
precisely estimated component.

## Boundary Violations

Boundary violations (estimates outside [0, 1]) were rare across
all scenarios, occurring in fewer than 1% of replicates. No
systematic pattern distinguished the two methods with respect to
boundary violations.

# Discussion

We find that both the Nakagawa-Schielzeth and Johnson methods
provide approximately unbiased estimates of marginal and
conditional $R^2$ for LMMs across a broad range of
data-generating conditions. The two
methods are mathematically identical for random-intercept
models, and this equivalence was confirmed empirically.
Differences emerged only in random-slope scenarios, where the
Johnson extension is theoretically preferred and showed modestly
lower bias.

Several patterns warrant attention. First, small samples
($J = 20$, $n = 5$) produced substantially wider sampling
distributions and lower monotonicity rates, suggesting that
$R^2$ estimates from small multilevel studies should be
interpreted with caution. Second, the interaction between ICC
and sample size had a notable effect on MSE: the marginal
$R^2$ was estimated most precisely in high-ICC, large-sample
conditions, where the variance decomposition is most
identifiable. Third, singular fits were common in low-ICC
random-slope scenarios, a well-known limitation of maximum
likelihood estimation for small variance components
[@batesLme4LinearMixed2015].

## Limitations and Future Directions

This study compared only two of the six available methods. The
Edwards et al. likelihood-ratio approach, the Jaeger et al.
semi-partial decomposition, the Xu level-specific framework,
and the `performance` package implementation are natural
candidates for an expanded comparison (plan-full.md); the
rationale for excluding each from the present empirical
comparison is stated in the "Summary of Methods" section above.
Such an expansion would increase the design to 324 scenarios
with 1,000 replications and, by extrapolation from the current
32-scenario/500-replicate run time, is *estimated* (not
benchmarked) to require approximately 5--6 hours of computation
on 8 cores; this figure has not been verified by a partial or
pilot run at the expanded scale and should be treated as a rough
projection only.

The random-slope conclusions in this study (Results,
Discussion) are conditional on a single, disclosed
slope-to-intercept variance ratio, $\tau_1 = 0.5\,\tau_0$ (see
Data-Generating Mechanisms), which was not varied factorially.
They should not be read as general claims across arbitrary
random-slope variance structures.

The fitted models use `lme4`'s default correlated
random-intercept/random-slope parameterization
(`(1 + x1 | cluster_id)`, an unstructured $G$ matrix), whereas the
data-generating process draws the random intercept and random
slope independently (`R/generate_data.R`). This mismatch reflects
standard applied practice -- most analysts fit an unstructured
$G$ matrix without first testing whether the true intercept-slope
correlation is zero -- but it means the fitted models estimate one
more variance-covariance parameter than the DGP requires, which
plausibly contributes to the elevated singular-fit rate observed
in low-ICC random-slope scenarios (Results, Convergence).

The present design is restricted to balanced clusters and
Gaussian outcomes. Extensions to unbalanced designs, binary or
count outcomes (GLMMs), and crossed random effects would
increase the practical relevance of the findings. The DGP
assumed independent predictors; correlated predictors could
affect the relative performance of methods that decompose
variance at the predictor level (Jaeger et al.) versus those
that operate on the total fixed-effect variance
(Nakagawa-Schielzeth, Johnson).

Finally, this study does not report coverage of confidence or
credible intervals for $\hat{R}^2$, a performance measure
recommended by @morrisUsingSimulationStudies2019 and standard in
$R^2$-estimator evaluations. It is omitted here because neither
`MuMIn::r.squaredGLMM()` nor the trigamma extension returns an
analytic standard error or interval for $R^2_m$/$R^2_c$;
constructing one would require a bootstrap or delta-method
addition to the estimation pipeline (`R/compute_r2.R`), which was
judged out of scope for the present two-method comparison. Adding
a bootstrap-based coverage measure is a concrete target for a
future revision of `R/performance_metrics.R`.

## Practical Recommendations

Based on these results, we offer the following guidance for
applied researchers:

1. For random-intercept models, the Nakagawa-Schielzeth and
   Johnson methods are interchangeable. Either may be reported.

2. For random-slope models, the Johnson extension (trigamma
   method) is preferred on theoretical grounds and showed
   marginally better empirical performance.

3. Both marginal and conditional $R^2$ should be reported, as
   they answer distinct questions about population-average
   versus cluster-specific prediction.

4. In small-sample settings ($J < 30$ or $n < 10$), $R^2$
   estimates should be accompanied by measures of uncertainty.
   Monte Carlo standard errors or bootstrap confidence intervals
   are appropriate for this purpose.

5. Convergence diagnostics should always be checked.
   Singular fits indicate that the random-effects structure may
   be over-parameterized for the available data, and $R^2$
   estimates from such fits should be interpreted cautiously.

# Conclusions

We have endeavored to provide the first systematic comparison
of the statistical properties of $R^2$ measures for linear
mixed-effects models under controlled conditions where the true
variance decomposition is known. Both the Nakagawa-Schielzeth
and Johnson methods performed well across a wide range of
scenarios, with the Johnson extension offering modest advantages
for random-slope models. In conclusion, the results underscore
the importance of adequate sample sizes for reliable $R^2$
estimation in multilevel designs and support the routine
reporting of both marginal and conditional $R^2$ alongside
convergence diagnostics.

# Conflict of interest

The author declares no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

The data and code that support the findings of this study are
openly available in the `mixedr2` project repository.
Per-replicate Monte Carlo outputs, ADEMP-compliant cell-level
summaries, simulation drivers, and the analysis-plan
documentation are versioned under
`analysis/data/derived_data/sim_results.rds`,
`analysis/scripts/`, and `docs/`. Specific paths are listed in
the Reproducibility section below.

# Reproducibility

This manuscript was rendered with R 4.4.x inside the
project's zzcollab Docker container (image hash recorded in
`Dockerfile`; package set captured in `renv.lock`). Principal
packages used by the simulation pipeline are `mixedr2` (the
project's R package implementing the ADEMP simulation
machinery), `lme4` (the underlying linear-mixed-effects
fitter), and `MuMIn` (both the Nakagawa-Schielzeth and Johnson
$R^2$ estimators, via `r.squaredGLMM()` with the default and
`method = "trigamma"` arguments respectively; see
`R/compute_r2.R`). The `performance` package is discussed in the
literature review as a comparator implementation of the same
statistical method but is not invoked anywhere in the simulation
pipeline that produced the reported numbers. Additional packages
are `furrr` and `future` (parallel processing), `dplyr`,
`tidyr`, and `ggplot2` (reporting), and `rmarkdown` plus
`bookdown` for the manuscript build.

**Random-number generator.** `analysis/scripts/simulation.R`
pins `RNGkind("L'Ecuyer-CMRG")` once at program start, before any
seeding, per Morris et al. (2019) Section 4.1. Per-replicate
seeds are not derived from a running counter; instead,
`create_scenario_grid()` (`R/generate_data.R`) calls
`set.seed(master_seed)` once, with `master_seed = 20260306`, and
then draws one seed per scenario-replicate via
`sample.int(.Machine$integer.max, total_reps)`, so every one of
the 16,000 replicate-level seeds is pre-generated and stored in
the scenario grid before any model fitting begins. Each
replicate's stored seed is consumed exactly once, by
`set.seed(seed)` inside `generate_lmm_data()`
(`R/generate_data.R`), immediately before that replicate's data
are drawn. Parallel execution across replicates within a
scenario uses `furrr::future_map_dfr()` with
`furrr_options(seed = NULL)` (`R/simulation.R::run_scenario()`):
`seed = NULL` is appropriate here, rather than a gap in the
seeding strategy, because RNG state is already made
reproducible explicitly by the pre-drawn, per-replicate
`set.seed()` call described above, so `furrr`'s own parallel-safe
seeding layer would be redundant. A reader who wants to reproduce
a specific replicate should regenerate the scenario grid with
`create_scenario_grid(master_seed = 20260306)` and read off that
replicate's `seed` column, not compute `master_seed + rep_idx`.

**Data and code availability.** Simulation outputs are
checkpointed at
`analysis/data/derived_data/sim_results.rds` (16{,}000
replicates across 32 scenarios). Driver scripts are at
`analysis/scripts/`. ADEMP planning documents are at
`docs/plan-short.md` and `docs/plan-full.md`. The R package
source is at `R/`. The manuscript source, bibliography
(`references.bib`), and SIM CSL file are at
`analysis/report/`.

# References

## Morris et al. (2019) ADEMP Compliance

We audited this simulation study against the reporting standards
proposed by Morris, White, and Crowther
[-@morrisUsingSimulationStudies2019]. This section reports the
audit re-run against the code that produced the results in this
manuscript, superseding the earlier
`docs/morris-audit-2026-04-17.md` snapshot, two of whose three
listed gaps are now closed in the code as it stands.

**Verdict:** Mostly compliant.

**Gaps closed since the prior audit:**

- `n_sim = 500` is now documented with an explicit Monte Carlo SE
  justification in `analysis/scripts/simulation.R` (bias MCSE
  $\le 0.002$ at empirical SE $\approx 0.04$ requires
  $n_{sim} \ge 400$; $n_{sim} = 500$ gives bias MCSE
  $\approx 0.0018$).
- `RNGkind("L'Ecuyer-CMRG")` is now pinned at the top of
  `analysis/scripts/simulation.R`, before any seeding.

**Gaps still open:**

- Post-run `.Random.seed` is not captured per replicate, so the
  exact RNG state at each replicate's start is reconstructible
  only by re-running `create_scenario_grid()` with the recorded
  master seed, not by inspecting a stored state directly.
- Coverage is not among the reported performance measures (see
  Limitations for the reason and a proposed remediation).

A remediation plan for the two open gaps is recorded in
`docs/morris-audit-2026-04-17.md` and will be taken up in a
subsequent revision.
