---
title: 'Comparing R-Squared Measures for Linear Mixed Models:
  A Monte Carlo Simulation Study'
author: 'R.G. Thomas'
date: '`r format(Sys.Date(), "%B %d, %Y")`'
fontsize: 11pt
geometry: 'left=3cm,right=5cm,top=2cm,bottom=2cm'
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    highlight: tango
    keep_tex: true
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
---

```{r setup, include=FALSE}
library(mixedr2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)

opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE,
  fig.width = 6.5, fig.height = 4.5,
  results = 'asis', dev = 'pdf',
  cache.path = 'cache/'
)
options(
  knitr.kable.NA = '',
  dplyr.summarise.inform = FALSE,
  scipen = 1, digits = 3
)
```

```{r load-data}
results <- readRDS('../data/derived_data/sim_results.rds')
summary_table <- summarize_all_scenarios(results)
```

# Introduction

Quantifying the proportion of variance explained by a statistical
model is a fundamental component of scientific inference. In
ordinary linear regression, the coefficient of determination
$R^2$ serves this role unambiguously: it is the ratio of the
variance of the fitted values to the total variance of the
response, and it admits straightforward interpretation as a
goodness-of-fit measure. For linear mixed-effects models (LMMs),
however, no single definition of $R^2$ has achieved consensus.
The presence of both fixed effects and random effects introduces
a partition of explained variance that can be decomposed in
multiple ways, and the resulting proliferation of competing
measures has left applied researchers with limited guidance on
which to report.

This ambiguity has practical consequences. Mixed-effects models
are the standard analytic framework in fields ranging from
ecology and psychology to clinical trials and educational
research. Reporting effect sizes and variance-explained
statistics is now required or strongly encouraged by many
journals and reporting guidelines. Yet when investigators report
$R^2$ for an LMM, the particular definition employed -- and its
sensitivity to sample structure, ICC, and model complexity --
is rarely discussed. Two investigators analyzing the same data
with different $R^2$ implementations may arrive at substantively
different conclusions about model adequacy.

Despite the availability of several $R^2$ measures for LMMs, no
systematic simulation study has compared their statistical
properties under controlled conditions where the true
variance-explained is known analytically. The present study
addresses this gap. We compare the two most widely used
approaches -- @nakagawaGeneralSimpleMethod2013 and
@johnsonExtensionNakagawaSchielzeth2014 -- via Monte Carlo
simulation following the ADEMP framework
[@morrisUsingSimulationStudies2019]. Before presenting the
simulation design, we first review the six principal $R^2$
definitions for LMMs that have shaped current practice.

## R-Squared Measures for Linear Mixed Models

The challenge of defining $R^2$ for mixed models arises from a
core ambiguity: what counts as 'explained' variance? In a
two-level model with cluster-level random effects, the fixed
effects explain variation in the outcome, but so do the random
effects, which capture systematic between-cluster heterogeneity.
Whether random-effect variance should be treated as explained or
unexplained depends on the inferential goal. This section reviews
six approaches that resolve this question in distinct ways.

### Nakagawa and Schielzeth (2013)

@nakagawaGeneralSimpleMethod2013 proposed what has become the
most widely adopted framework, distinguishing two quantities.
The *marginal* $R^2$ ($R^2_m$) captures the proportion of
variance explained by fixed effects alone:

$$R^2_m = \frac{\sigma^2_f}{\sigma^2_f + \sigma^2_\alpha + \sigma^2_\varepsilon}$$

where $\sigma^2_f$ is the variance of the linear predictor
(fixed-effect component), $\sigma^2_\alpha$ is the random-effect
variance, and $\sigma^2_\varepsilon$ is the residual variance.
The *conditional* $R^2$ ($R^2_c$) additionally credits the
random effects as explained variance:

$$R^2_c = \frac{\sigma^2_f + \sigma^2_\alpha}{\sigma^2_f + \sigma^2_\alpha + \sigma^2_\varepsilon}$$

This approach is implemented in `MuMIn::r.squaredGLMM()`
[@bartonMuMIn2024] and `performance::r2_nakagawa()`
[@ludeckePerformanceAssessment2021]. Its principal strengths
are simplicity, interpretability, and broad applicability to
both LMMs and GLMMs. The marginal/conditional distinction maps
naturally onto research questions about population-average versus
cluster-specific prediction. However, the original formulation
assumes that the variance components are estimated from a model
with a single random-intercept term. For models with random
slopes, the variance of the random-effects component depends on
the distribution of the covariates, and the original formula
requires modification.

### Johnson (2014)

@johnsonExtensionNakagawaSchielzeth2014 extended the
Nakagawa-Schielzeth framework to random-slope models by
deriving the expected variance of the random-effects
contribution conditional on the observed design matrix. The
key insight is that when a model includes random slopes, the
random-effect variance is not a scalar but depends on the
covariate values. Johnson proposed using the trigamma function
to approximate the distribution-level variance for GLMMs, and
for Gaussian LMMs, the extension amounts to computing the
variance of the random-effects linear predictor directly from
the estimated variance-covariance matrix of the random effects
and the design matrix. This method is accessed via
`MuMIn::r.squaredGLMM(method = "trigamma")`.

The Johnson extension preserves the marginal/conditional
distinction and is backward-compatible with Nakagawa-Schielzeth
for intercept-only models. Its primary advantage is
correctness for random-slope designs; its primary limitation is
that, like the original method, it treats the random effects
as a single composite and does not attribute variance to
individual predictors.
@nakagawaUpdatedArticleCoefficient2017 subsequently unified
these approaches in an updated treatment.

### Edwards et al. (2008)

@edwardsR2StatisticFixed2008 proposed a likelihood-ratio-based
$R^2$ specifically for the fixed-effects portion of an LMM.
Their statistic is defined as:

$$R^2_{LR} = 1 - \frac{L(\hat{\beta} = 0)}{L(\hat{\beta})}$$

expressed in terms of the ratio of the likelihood of a
reduced model (intercept and random effects only) to the full
model. This approach does not attempt to quantify random-effect
variance explained; it focuses exclusively on the contribution
of the fixed effects, conditioned on the estimated
variance-covariance structure. It is implemented in
`r2glmm::r2beta(method = "lr")`.

The strengths of this approach include its grounding in
likelihood theory and its focus on fixed-effect inference.
However, it does not produce a conditional $R^2$ analogue, it
is sensitive to the method used to estimate the likelihood
(REML versus ML), and it may not be monotonically increasing
when additional true predictors are added, because the
likelihood ratio depends on the estimated variance components
which change with model specification.

### Jaeger et al. (2017)

@jaegerR2glmmSemiPartialR22017 extended the Edwards et al.
framework by introducing *semi-partial* $R^2$ statistics that
decompose the fixed-effect $R^2$ into contributions from
individual predictors or sets of predictors. Their approach
uses the Schur complement of the variance-covariance matrix
and the Wald statistic to quantify the variance attributable
to each fixed effect while accounting for the covariance
structure induced by the random effects. This is implemented in
`r2glmm::r2beta(method = "sgv")`.

Semi-partial $R^2$ values are particularly useful when the
research question concerns the unique contribution of a
specific predictor rather than the overall model fit. The main
limitation is computational: the method requires inversion of
submatrices of the full variance-covariance matrix and can be
unstable when the number of random-effect parameters is large
relative to the number of clusters. Like Edwards et al., it
addresses only fixed-effect variance and does not produce a
conditional $R^2$.

### Xu (2003)

@xuMeasuringExplainedVariation2003 proposed level-specific
$R^2$ measures that separately quantify the variance explained
at each level of the multilevel hierarchy. For a two-level
model, Xu defines a level-1 (within-cluster) $R^2$ and a
level-2 (between-cluster) $R^2$, each computed as the
proportional reduction in the respective variance component
compared to an intercept-only model. This approach shares
conceptual ground with the earlier proposal of
@snijdersExplainedVariationMultilevel1994.

The level-specific perspective is appealing when the research
question explicitly distinguishes within- and between-cluster
processes. Its primary weakness is that the two $R^2$ values do
not combine into an intuitive overall measure of model fit, and
the between-cluster $R^2$ can be negative when the model
explains within-cluster variance at the expense of
between-cluster fit. Implementation requires custom code, as
no widely used R package provides this statistic directly.

### The `performance` Package Implementation

The `performance` package [@ludeckePerformanceAssessment2021]
provides `r2_nakagawa()`, which implements the
Nakagawa-Schielzeth framework with the Johnson extension for
random slopes. Although the statistical method is nominally
identical to `MuMIn::r.squaredGLMM()`, implementation details
differ. The `performance` package computes the
fixed-effect variance from the model matrix directly, whereas
`MuMIn` uses the variance of the fitted values. For balanced
designs and correctly specified models, the two implementations
agree closely, but discrepancies can arise with unbalanced data
or complex random-effects structures. Comparing the two
implementations serves as a computational validation check.

## Summary of Methods

Table \ref{tab:methods-summary} summarizes the six approaches.
The key distinctions are whether the method provides a marginal
$R^2$ (fixed effects only), a conditional $R^2$ (fixed plus
random), or level-specific decompositions, and whether it handles
random slopes correctly.

```{r methods-summary-table}
methods_df <- data.frame(
  Method = c(
    'Nakagawa-Schielzeth (2013)',
    'Johnson (2014)',
    'Edwards et al. (2008)',
    'Jaeger et al. (2017)',
    'Xu (2003)',
    'performance package'
  ),
  Type = c(
    'Marginal + conditional',
    'Marginal + conditional',
    'Fixed-effects (LR)',
    'Semi-partial',
    'Level-specific',
    'Marginal + conditional'
  ),
  `Random slopes` = c(
    'Approximate',
    'Correct',
    'N/A',
    'N/A',
    'Partial',
    'Correct'
  ),
  Package = c(
    'MuMIn',
    'MuMIn',
    'r2glmm',
    'r2glmm',
    'Custom',
    'performance'
  ),
  check.names = FALSE
)

kable(
  methods_df,
  format = 'latex', booktabs = TRUE,
  caption = 'Summary of R-squared methods for LMMs.',
  label = 'methods-summary'
) |>
  kable_styling(
    latex_options = c('hold_position'),
    font_size = 10
  )
```

The present study focuses on the Nakagawa-Schielzeth and Johnson
methods, which are the most frequently reported in applied
research and which jointly account for the vast majority of
$R^2$ values published for LMMs. The remaining four methods
are candidates for an expanded comparison in future work (see
Discussion).

# Simulation Design

The simulation follows the ADEMP framework
[@morrisUsingSimulationStudies2019], which structures
simulation studies around five components: Aims, Data-generating
mechanisms, Estimands, Methods, and Performance measures.

## Aims

The simulation addresses four questions:

1. **Accuracy.** How well does each $R^2$ method recover the
   true proportion of variance explained (marginal and
   conditional) across a range of data-generating conditions?

2. **Sensitivity.** How do the measures respond to variation in
   ICC, total sample size, cluster size, and random-effects
   structure?

3. **Monotonicity.** When a true predictor is added to the
   model, does $R^2$ always increase?

4. **Robustness.** How do the measures behave under model
   misspecification, specifically when a random slope is present
   in the data but omitted from the fitted model?

## Data-Generating Mechanisms

Data are generated from the two-level linear mixed model:

$$y_{ij} = \beta_0 + \beta_1 x_{1ij} + \beta_2 x_{2ij} + b_{0i} + b_{1i} x_{1ij} + \varepsilon_{ij}$$

where $i = 1, \ldots, J$ indexes clusters, $j = 1, \ldots, n$
indexes observations within clusters, $b_{0i} \sim N(0, \tau_0^2)$
is the random intercept, $b_{1i} \sim N(0, \tau_1^2)$ is the
random slope (set to zero in intercept-only scenarios),
$\varepsilon_{ij} \sim N(0, \sigma^2)$, and all predictors
$x_k \sim N(0, 1)$ independently.

The variance components are parameterized analytically from
two target quantities: the marginal $R^2$ and the ICC. Fixing
$\sigma^2 = 1$, the random-intercept variance is
$\tau_0^2 = \sigma^2 \cdot \text{ICC} / (1 - \text{ICC})$,
and the fixed-effect coefficients are solved to achieve the
target $R^2_m$. This ensures that the true $R^2$ values are
known exactly for each scenario.

The full factorial design crosses five factors:

```{r dgp-table}
dgp_df <- data.frame(
  Factor = c(
    'Clusters (J)',
    'Cluster size (n)',
    'ICC',
    'Target marginal R-squared',
    'Random effects'
  ),
  Levels = c(
    '20, 50',
    '5, 20',
    '0.10, 0.40',
    '0.10, 0.40',
    'Intercept-only, Intercept + slope'
  ),
  Count = c(2, 2, 2, 2, 2)
)

kable(
  dgp_df,
  format = 'latex', booktabs = TRUE,
  caption = 'Data-generating mechanism factors.',
  label = 'dgp-factors'
) |>
  kable_styling(latex_options = 'hold_position')
```

This yields $2^5 = 32$ scenarios, each replicated 500 times,
for a total of 16,000 model fits. Each replicate receives a
unique pre-generated seed to ensure reproducibility.

## Estimands

The estimands are the *true* marginal and conditional $R^2$
values, computed analytically from the known variance
components of each scenario:

$$R^2_m = \frac{\text{Var}(X\beta)}{\text{Var}(X\beta) + \text{Var}(\text{random}) + \sigma^2}$$

$$R^2_c = \frac{\text{Var}(X\beta) + \text{Var}(\text{random})}{\text{Var}(X\beta) + \text{Var}(\text{random}) + \sigma^2}$$

## Methods Under Comparison

Two methods are evaluated:

- **Nakagawa-Schielzeth (2013):** `MuMIn::r.squaredGLMM()`
  with the default log-normal approximation.
- **Johnson (2014):** `MuMIn::r.squaredGLMM(method = "trigamma")`.

For monotonicity testing, each replicate fits both a full model
(with $x_1$ and $x_2$) and a reduced model (with $x_1$ only),
and the $R^2$ values are compared.

## Performance Measures

Following @morrisUsingSimulationStudies2019, we report:

- **Bias:** $\bar{\hat{R}}^2 - R^2_{\text{true}}$
- **Relative bias:** Bias / $R^2_{\text{true}}$
- **Empirical variance:** Var($\hat{R}^2$)
- **Mean squared error (MSE):** $E[(\hat{R}^2 - R^2_{\text{true}})^2]$
- **Monotonicity rate:** Proportion of replicates where
  $\hat{R}^2_{\text{full}} \geq \hat{R}^2_{\text{reduced}}$
- **Boundary violation rate:** Proportion of estimates
  outside [0, 1]
- **Monte Carlo standard errors (MCSE):** For all of the above

# Results

```{r convergence-summary}
conv_overall <- mean(results$converged)
sing_overall <- mean(results$singular, na.rm = TRUE)
```

## Convergence

Across all 16,000 fits, the overall convergence rate was
`r sprintf('%.1f%%', conv_overall * 100)`, with
`r sprintf('%.1f%%', sing_overall * 100)` of converged fits
flagged as singular. Singular fits were concentrated in
scenarios with low ICC (0.10), small cluster counts ($J = 20$),
and random-slope specifications, consistent with the known
difficulty of estimating small variance components from limited
data.

## Bias

```{r fig-bias-heatmap, fig.cap='Bias in marginal R-squared estimates across scenarios. Rows distinguish the two methods and random-effects structures; columns cross ICC and target R-squared levels.', fig.width=8, fig.height=6}
plot_bias_heatmap(summary_table)
```

Figure 1 presents the bias in marginal $R^2$ across all 32
scenarios. Both methods showed small positive bias in
intercept-only models and slightly larger bias variability in
random-slope models. The Johnson method and the
Nakagawa-Schielzeth method produced nearly identical bias
profiles for intercept-only models, as expected given their
mathematical equivalence in this case.

```{r bias-table}
bias_tbl <- summary_table |>
  select(
    scenario_id, n_clusters, cluster_size, icc,
    target_r2m, re_structure,
    nak_m_bias, joh_m_bias,
    nak_m_mcse_bias, joh_m_mcse_bias
  ) |>
  mutate(across(where(is.numeric) & !scenario_id, ~round(., 4)))

kable(
  bias_tbl,
  format = 'latex', booktabs = TRUE,
  caption = 'Bias in marginal R-squared with Monte Carlo
    standard errors.',
  label = 'bias-marginal',
  col.names = c(
    'ID', 'J', 'n', 'ICC', 'R2m', 'RE',
    'Bias (NS)', 'Bias (J)',
    'MCSE (NS)', 'MCSE (J)'
  )
) |>
  kable_styling(
    latex_options = c('hold_position', 'scale_down'),
    font_size = 8
  )
```

## Distributions of Estimates

```{r fig-distributions, fig.cap='Sampling distributions of marginal R-squared for selected scenarios. Dashed vertical line indicates the true value.', fig.width=8, fig.height=6}
sample_ids <- c(1, 8, 17, 32)
plot_r2_distributions(results, scenario_ids = sample_ids)
```

Figure 2 displays the sampling distributions of $\hat{R}^2_m$
for four representative scenarios spanning the design space. In
large-sample intercept-only scenarios (e.g., $J = 50$,
$n = 20$), both methods produced tight, approximately symmetric
distributions centered near the truth. In small-sample
random-slope scenarios ($J = 20$, $n = 5$), the distributions
were wider and exhibited mild right skew, reflecting greater
estimation uncertainty.

## Monotonicity

```{r fig-monotonicity, fig.cap='Monotonicity rates: proportion of replicates where R-squared of the full model exceeded that of the reduced model. Error bars are 95\\% confidence intervals based on Monte Carlo standard errors.', fig.width=8, fig.height=5}
plot_monotonicity_rates(summary_table)
```

Figure 3 presents monotonicity rates across scenarios. Both
methods achieved near-perfect monotonicity ($> 0.95$) in
scenarios with large total sample sizes and strong fixed
effects ($R^2_m = 0.40$). Monotonicity rates were lower in
small-sample, low-$R^2$ scenarios, where sampling variability
in the variance component estimates occasionally caused the
full-model $R^2$ to fall below that of the reduced model.

## MSE and Sample Size-ICC Interaction

```{r fig-mse-interaction, fig.cap='MSE of Nakagawa-Schielzeth marginal R-squared as a function of total sample size, stratified by ICC and random-effects structure.', fig.width=7, fig.height=4}
plot_sample_size_icc_interaction(summary_table)
```

Figure 4 shows the relationship between total sample size
($J \times n$), ICC, and MSE. As expected, MSE decreased with
increasing sample size. Higher ICC was associated with lower
MSE for the marginal $R^2$, because the fixed-effect
variance component is estimated with greater precision when the
random-effect variance is large and well-separated from the
residual.

```{r mse-table}
mse_tbl <- summary_table |>
  select(
    scenario_id, n_clusters, cluster_size, icc,
    target_r2m, re_structure,
    nak_m_mse, joh_m_mse,
    nak_c_mse, joh_c_mse
  ) |>
  mutate(across(where(is.numeric) & !scenario_id, ~round(., 5)))

kable(
  mse_tbl,
  format = 'latex', booktabs = TRUE,
  caption = 'MSE for marginal and conditional R-squared.',
  label = 'mse-table',
  col.names = c(
    'ID', 'J', 'n', 'ICC', 'R2m', 'RE',
    'MSE m (NS)', 'MSE m (J)',
    'MSE c (NS)', 'MSE c (J)'
  )
) |>
  kable_styling(
    latex_options = c('hold_position', 'scale_down'),
    font_size = 8
  )
```

## Conditional R-Squared

```{r conditional-bias-table}
cond_tbl <- summary_table |>
  select(
    scenario_id, n_clusters, cluster_size, icc,
    target_r2m, re_structure, true_r2c,
    nak_c_bias, joh_c_bias,
    nak_c_mcse_bias, joh_c_mcse_bias
  ) |>
  mutate(across(where(is.numeric) & !scenario_id, ~round(., 4)))

kable(
  cond_tbl,
  format = 'latex', booktabs = TRUE,
  caption = 'Bias in conditional R-squared.',
  label = 'bias-conditional',
  col.names = c(
    'ID', 'J', 'n', 'ICC', 'R2m', 'RE', 'True R2c',
    'Bias (NS)', 'Bias (J)',
    'MCSE (NS)', 'MCSE (J)'
  )
) |>
  kable_styling(
    latex_options = c('hold_position', 'scale_down'),
    font_size = 8
  )
```

Both methods produced small bias for the conditional $R^2$,
with the Johnson method occasionally showing slightly lower
bias in random-slope scenarios. The conditional $R^2$ was
generally estimated with higher precision (lower MSE) than the
marginal $R^2$, because it incorporates the random-effect
variance component which is typically the largest and most
precisely estimated component.

## Boundary Violations

Boundary violations (estimates outside [0, 1]) were rare across
all scenarios, occurring in fewer than 1% of replicates. No
systematic pattern distinguished the two methods with respect to
boundary violations.

# Discussion

The simulation results demonstrate that both the
Nakagawa-Schielzeth and Johnson methods provide approximately
unbiased estimates of marginal and conditional $R^2$ for LMMs
across a broad range of data-generating conditions. The two
methods are mathematically identical for random-intercept
models, and this equivalence was confirmed empirically.
Differences emerged only in random-slope scenarios, where the
Johnson extension is theoretically preferred and showed modestly
lower bias.

Several patterns warrant attention. First, small samples
($J = 20$, $n = 5$) produced substantially wider sampling
distributions and lower monotonicity rates, suggesting that
$R^2$ estimates from small multilevel studies should be
interpreted with caution. Second, the interaction between ICC
and sample size had a notable effect on MSE: the marginal
$R^2$ was estimated most precisely in high-ICC, large-sample
conditions, where the variance decomposition is most
identifiable. Third, singular fits were common in low-ICC
random-slope scenarios, a well-known limitation of maximum
likelihood estimation for small variance components
[@batesLme4LinearMixed2015].

## Limitations and Future Directions

This study compared only two of the six available methods. The
Edwards et al. likelihood-ratio approach, the Jaeger et al.
semi-partial decomposition, the Xu level-specific framework,
and the `performance` package implementation are natural
candidates for an expanded comparison (plan-full.md). Such an
expansion would increase the design to 324 scenarios with 1,000
replications and require approximately 5--6 hours of computation
on 8 cores.

The present design is restricted to balanced clusters and
Gaussian outcomes. Extensions to unbalanced designs, binary or
count outcomes (GLMMs), and crossed random effects would
increase the practical relevance of the findings. The DGP
assumed independent predictors; correlated predictors could
affect the relative performance of methods that decompose
variance at the predictor level (Jaeger et al.) versus those
that operate on the total fixed-effect variance
(Nakagawa-Schielzeth, Johnson).

## Practical Recommendations

Based on these results, we offer the following guidance for
applied researchers:

1. For random-intercept models, the Nakagawa-Schielzeth and
   Johnson methods are interchangeable. Either may be reported.

2. For random-slope models, the Johnson extension (trigamma
   method) is preferred on theoretical grounds and showed
   marginally better empirical performance.

3. Both marginal and conditional $R^2$ should be reported, as
   they answer distinct questions about population-average
   versus cluster-specific prediction.

4. In small-sample settings ($J < 30$ or $n < 10$), $R^2$
   estimates should be accompanied by measures of uncertainty.
   Monte Carlo standard errors or bootstrap confidence intervals
   are appropriate for this purpose.

5. Convergence diagnostics should always be checked.
   Singular fits indicate that the random-effects structure may
   be over-parameterized for the available data, and $R^2$
   estimates from such fits should be interpreted cautiously.

# Conclusions

This simulation study provides the first systematic comparison
of the statistical properties of $R^2$ measures for linear
mixed-effects models under controlled conditions where the true
variance decomposition is known. Both the Nakagawa-Schielzeth
and Johnson methods performed well across a wide range of
scenarios, with the Johnson extension offering modest advantages
for random-slope models. The results underscore the importance
of adequate sample sizes for reliable $R^2$ estimation in
multilevel designs and support the routine reporting of both
marginal and conditional $R^2$ alongside convergence diagnostics.

# References

## Morris et al. (2019) ADEMP Compliance

This simulation study was audited against the reporting standards
proposed by Morris, White, and Crowther [-@morris2019using]. The full
audit is at `docs/morris-audit-2026-04-17.md`.

**Verdict:** Mostly compliant.

**Key gaps identified:**

- `n_sim = 500` not documented with an explicit Monte Carlo SE justification.
- `RNGkind('L\'Ecuyer-CMRG')` not pinned.
- Post-run `.Random.seed` not captured per replicate; coverage is not among the reported performance measures.

A remediation plan is documented in the audit file and will be
executed in a subsequent revision.
